How to select predictive models for decision-making or causal inference?

### 2023-12-17 how_to_select_causal_models.tex
iter

Indeed, concluding on the effect of a treatment is a difficult
causal-inference task, as it can be easily compromised by confounding:
spurious associations between treatment
allocation and baseline health, \eg only prescribing a drug to mild cases
\citep{hernan_causal_2020,vanderweele2019principles}.
Predictive modeling bridges to causal inference theory under the name of
\emph{outcome models} (or G-computation, 
G-formula \cite{robins_role_1986}, Q-model
\cite{snowden_implementation_2011}, conditional mean regression
\cite{wendling_comparing_2018}). 
% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
% alleviate this concern. Yet most machine-learning models are trained on
% \emph{observational} data, close to real-word practice
% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
Medical statistics and epidemiology have mostly used other
causal-inference methods, modeling treatment assignment
with propensity scores \cite{rosenbaum_central_1983,austin_moving_2015,casucci2018estimating,grose_use_2020}. Outcome modeling brings
the benefit of going
beyond average effects, estimating individualized or conditional average
treatment effects (CATE), central to precision medicine.
%
For this purpose, such methods are also invaluable on randomized trials
\cite{su2018random,lamont2018identification,hoogland2021tutorial}.

@@

We first illustrate the problem of causal model selection and briefly review
prior art. Then we anchor causal model selection in the \emph{potential outcome}
framework and details the causal risks and model-selection procedure. We then
rewrite the so-called $R\text{-risk}$ as a reweighted version of mean squared
difference between the true and estimated individualized treatment effect.
Finally, we conduct a thorough empirical study comparing the different metrics
on diverse datasets, using a family of simulations and real health
data, going beyond prior work limited to specific simulation settings
\citep{schuler_comparison_2018,alaa_validating_2019} (Appendix
\ref{apd:prior_work}).

@@

Using a predictor to reason on causal effects relies on contrasting the
prediction of the outcome for a given individual with and without the treatment.
%
Given various predictors of the outcome, which one should we use?
%
Standard predictive modeling or machine-learning practice selects the predictor
that minimizes the expected error on the outcome
\citep{poldrack2020establishment,varoquaux2022evaluating}. However, this
predictor may not be the best model to reason about causal effects of an
intervention as Figure \ref{fig:toy_example} illustrates. Consider the
probability $Y$ of an undesirable outcome (\eg death), a binary treatment $A \in
    \{0, 1\}$, and a covariate $X \in \mathbb R$ summarizing the patient health
status (\eg the Charlson index \cite{charlson_new_1987}). We simulate a
treatment beneficial (decreases mortality) for patients with high Charlson
scores (bad health status) but with little effect for patients in good condition
(low Charlson scores).

@@

Beyond rankings, the differences in terms of absolute
ability to select the best model are large: The R-risk selects a model
with a $\tau\text{-risk}$ only 1\% higher
than the best
possible candidate for strong overlap on Caussim, but selecting with
the $\mu\text{-risk}$ or $\mu\text{-risk}_{IPW}$ --as per machine-learning
practice-- leads to 10\% excess risk and using $\tau\text{-risk}_{IPW}$
--as in some causal-inference methods \citep{athey2016recursive,gutierrez_causal_2016}--leads to 100\% excess risk
(Figure
\ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}). Across
datasets, the $R\text{-risk}$ consistently decreases the
risk compared to the $\mu\text{-risk}$: from
0.1\% to 1\% on ACIC2016,  1\% from to 20\% on ACIC2018,
and 0.05\% from to 1\% on Twins.

@@

Several theoretical works have proposed causal model selection procedures
that are \emph{consistent}: select the best model in a family given
asymptotically large data. These work rely on introducing a
CATE estimator in the testing procedure: matching
\citep{rolling_model_2014}, an IPW estimate
\citep{gutierrez_causal_2016}, a doubly robust estimator
\citep{saito_counterfactual_2020}, or debiasing the error with influence
functions \cite{alaa_validating_2019}. However, for theoretical
guarantees to hold, the test-set correction needs to converge to the
oracle: it needs to be flexible enough --well-posed-- and asymptotic
data. From a practical perspective, meeting such requirements
implies having a good CATE estimate, thus having solved
the original problem of causal model selection.

>>>>

Indeed, concluding on the effect of a treatment is a difficult
causal-inference task, as it can be easily compromised by confounding:
spurious associations between treatment
allocation and baseline health, \eg only prescribing a drug to mild cases
\cite{hernan_causal_2020,vanderweele2019principles}.
Predictive modeling bridges to causal inference theory under the name of
\emph{outcome models} (or G-computation, 
G-formula \cite{robins_role_1986}, Q-model
\cite{snowden_implementation_2011}, conditional mean regression
\cite{wendling_comparing_2018}). 
% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
% alleviate this concern. Yet most machine-learning models are trained on
% \emph{observational} data, close to real-word practice
% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
Medical statistics and epidemiology have mostly used other
causal-inference methods, modeling treatment assignment
with propensity scores \cite{rosenbaum_central_1983,austin_moving_2015,casucci2018estimating,grose_use_2020}. Outcome modeling brings
the benefit of going
beyond average effects, estimating individualized or conditional average
treatment effects (CATE), central to precision medicine.
%
For this purpose, such methods are also invaluable on randomized trials
\cite{su2018random,lamont2018identification,hoogland2021tutorial}.

@@

We first illustrate the problem of causal model selection and briefly review
prior art. Then we anchor causal model selection in the \emph{potential outcome}
framework and details the causal risks and model-selection procedure. We then
rewrite the so-called $R\text{-risk}$ as a reweighted version of mean squared
difference between the true and estimated individualized treatment effect.
Finally, we conduct a thorough empirical study comparing the different metrics
on diverse datasets, using a family of simulations and real health
data, going beyond prior work limited to specific simulation settings
\cite{schuler_comparison_2018,alaa_validating_2019} (Appendix
\ref{apd:prior_work}).

@@

Using a predictor to reason on causal effects relies on contrasting the
prediction of the outcome for a given individual with and without the treatment.
%
Given various predictors of the outcome, which one should we use?
%
Standard predictive modeling or machine-learning practice selects the predictor
that minimizes the expected error on the outcome
\cite{poldrack2020establishment,varoquaux2022evaluating}. However, this
predictor may not be the best model to reason about causal effects of an
intervention as Figure \ref{fig:toy_example} illustrates. Consider the
probability $Y$ of an undesirable outcome (\eg death), a binary treatment $A \in
    \{0, 1\}$, and a covariate $X \in \mathbb R$ summarizing the patient health
status (\eg the Charlson index \cite{charlson_new_1987}). We simulate a
treatment beneficial (decreases mortality) for patients with high Charlson
scores (bad health status) but with little effect for patients in good condition
(low Charlson scores).

@@

Beyond rankings, the differences in terms of absolute
ability to select the best model are large: The R-risk selects a model
with a $\tau\text{-risk}$ only 1\% higher
than the best
possible candidate for strong overlap on Caussim, but selecting with
the $\mu\text{-risk}$ or $\mu\text{-risk}_{IPW}$ --as per machine-learning
practice-- leads to 10\% excess risk and using $\tau\text{-risk}_{IPW}$
--as in some causal-inference methods \cite{athey2016recursive,gutierrez_causal_2016}--leads to 100\% excess risk
(Figure
\ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}). Across
datasets, the $R\text{-risk}$ consistently decreases the
risk compared to the $\mu\text{-risk}$: from
0.1\% to 1\% on ACIC2016,  1\% from to 20\% on ACIC2018,
and 0.05\% from to 1\% on Twins.

@@

Several theoretical works have proposed causal model selection procedures
that are \emph{consistent}: select the best model in a family given
asymptotically large data. These work rely on introducing a
CATE estimator in the testing procedure: matching
\cite{rolling_model_2014}, an IPW estimate
\cite{gutierrez_causal_2016}, a doubly robust estimator
\cite{saito_counterfactual_2020}, or debiasing the error with influence
functions \cite{alaa_validating_2019}. However, for theoretical
guarantees to hold, the test-set correction needs to converge to the
oracle: it needs to be flexible enough --well-posed-- and asymptotic
data. From a practical perspective, meeting such requirements
implies having a good CATE estimate, thus having solved
the original problem of causal model selection.
